Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-04-23 paper/main.tex
move figure 1?
working on section 2

\documentclass[11pt]{article}
\usepackage{hyperref}
\usepackage{amsmath, amsfonts, amssymb, mathrsfs, dsfont}
\usepackage{dcolumn}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{filemod}
\usepackage{natbib}
\usepackage[ruled, vlined]{algorithm2e}
\usepackage{floatrow}
\usepackage{setspace}
\usepackage{verbatim}
\usepackage{graphicx}

@@

The Mira-Titan simulation suite \citep{moran2023mira}
defines ``cosmologies'' according to eight cosmological parameters (more on this
in Section \ref{sec:data}).  For a particular cosmology, the simulation of the power matter spectra 
is rather complex.  Although we presume there exists some true matter spectra as a function
of wavenumber, we are not able to directly observe it.  Rather, the simulation outputs 
a convoluted patchwork of potential spectra, which vary in fidelity and which are trusted
only over particular wavenumber ranges.  Our first objective is to synthesize expert knowledge
and all available simulation information into an effective estimate of the true underlying
matter spectrum.  \textcolor{red}{(In this paragraph, we need to narrate Figure 1.  The
figure provides a useful visual, but it needs to be talked through.  Comment on how
the weighted average takes into account the appropriate ranges.  Explain why the weighted
average on its own is not an effective estimate.  Motivate why we need a more complex model.
This is also a good place to introduce wavenumber as $k$.  Is there a particular reason
we plot it on the log scale?)}

@@

The remainder of the paper is organized as follows.  Section \ref{sec:data} 
introduces our motivating application, the Mira-Titan simulation suite.  
Section \ref{sec:hm_fit} describes our hierarchical Bayesian DGP and 
validates its performance on synthetic exercises and the CAMB simulations.  
Leveraging this trained model, Section \ref{sec:pred} details 
our procedure for predicting at unobserved cosmologies, benchmarking against
state-of-the-art competitors on the CAMB and Mira-Titan simulations. 
Section \ref{sec:disc} concludes with a discussion of our contributions and avenues for 
future research.  We provide reproducible code and an {\sf R} package for our 
Bayesian DGP model in a public git repository.\footnote{\url{https://github.com/stevewalsh124/dgp.hm}}
\textcolor{red}{Please note: this repository is currently private until we 
hear back regarding permissions for sharing the Mira-Titan data alongside our code.}

@@

\section{Data}
\label{sec:data}

The Mira-Titan simulation suite is a dataset consisting of simulated power spectra for 
117 different cosmologies. Each of these cosmologies is specified by eight cosmological 
parameter values: matter density, baryon density, amplitude of density fluctuations, 
dimensionless Hubble parameter, spectral index of scalar perturbations,
dark energy equation of state parameters, and neutrino density. For more details 
on these parameters and their effects, see \cite{dodelson2020modern, aghanim2020planck, heitmann2016mira}. 
Parameter ranges for each of these parameters are specified in Table 2.1 of \cite{moran2023mira}.
For simplicity, we will denote these eight parameters as $\theta \in \mathbb{R}^8$. Different simulation 
output exists for different values of a ninth value, the redshift; in this work, we fix 
this value at 0. For each cosmology, we have a batch of 18 simulated spectra: an inexpensive 
spectrum estimated from linear perturbation theory ($y_p$), 16 spectra estimated from 
lower-resolutions simulations ($y_{\ell_r}, r \in \{1,\dots,16\}$), and one spectrum from 
a high resolution simulation ($y_h$).

These spectra are represented as a function of wavenumber $k$, and hence is denoted $P(k)$. 
Following \cite{moran2023mira}, we model on the emulation space 
$\mathcal{P}(k)=\log_{10}\left(\frac{k^{1.5}P(k)}{2\pi^2}\right)$. The output is available 
for $0.001 \leq k \leq 5$ for $n=351$ different values of $k$, but each data type has differing 
values where it is deemed approximately unbiased. For $k<0.04$, only $y_p$ provides a reliable 
estimate of the spectra. $y_{\ell_r}$ are valid for $0.04 \leq k \leq 0.25$, and $y_h$ is valid 
for $0.04 \leq k \leq 5$. Figure \ref{fig:plot_data} shows an example of the output for a 
particular cosmology.

From the multiple low- and high-resolution spectra across all cosmologies, 
\cite{moran2023mira} obtained precision estimates of these spectra across the $n$ 
wavenumber values ($p_1,\dots,p_n$) using a log-log regression model, and a multiplier 
$c$ for the increase in precision from the low- to high- resolution output. Given 
that the perturbation theory spectra is non-stochastic, we treat this as having a 
precision of $10^5$. We leverage these precision estimates within our model at each 
data type's unbiased wavenumber values to define 
$\Lambda_p = \text{diag}\left((10^5)_{k < 0.04}\right)$, 
$\Lambda_{\ell} = \text{diag}\left(16(p_1,\dots,p_n)_{0.04 \leq k < 0.25}\right)$, 
$\Lambda_h = \text{diag}\left(c(p_1,\dots,p_n)_{0.04 \leq k \leq 5}\right)$. 
Here, each precision matrix is diagonal and the diagonal elements consist of the 
precision values or zero, depending on whether the $k$ value is approximately unbiased 
for that data type. With this, we calculate a weighted average for each cosmology: 
$\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_h y_h + \Lambda_{\ell} \bar{y}_\ell)$, 
where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$ is the precision matrix for 
the weighted average for a particular cosmology.

We will handle the data in two stages.  First, we will focus on 111 training cosmologies 
and obtain the best estimated spectrum for each.  Second, we will use these estimates 
to predict the spectra for six held-out cosmologies.

>>>>

\documentclass[11pt]{article}
\usepackage{hyperref}
\usepackage{amsmath, amsfonts, amssymb, mathrsfs, dsfont}
\usepackage{dcolumn}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{filemod}
\usepackage{natbib}
\usepackage[ruled, vlined]{algorithm2e}
\usepackage{floatrow}
\usepackage{setspace}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{xcolor} % only for coloring text while revising

@@

The Mira-Titan simulation suite \citep{moran2023mira}
defines ``cosmologies'' according to eight cosmological parameters (more on this
in Section \ref{sec:data}).  For a particular cosmology, the simulation of the power matter spectra 
is rather complex.  Although we presume there exists some true matter spectra as a function
of wavenumber, we are not able to directly observe it.  Rather, the simulation outputs 
a convoluted patchwork of potential spectra, which vary in fidelity and which are trusted
only over particular wavenumber ranges.  Our first objective is to synthesize expert knowledge
and all available simulation information into an effective estimate of the true underlying
matter spectrum.

@@

The remainder of the paper is organized as follows.  Section \ref{sec:data} 
introduces our motivating application, the Mira-Titan simulation suite.  
Section \ref{sec:hm_fit} describes our hierarchical Bayesian DGP and 
validates its performance on synthetic exercises and the CAMB simulations.  
Leveraging this trained model, Section \ref{sec:pred} details 
our procedure for predicting at unobserved cosmologies, benchmarking against
state-of-the-art competitors on the CAMB and Mira-Titan simulations. 
Section \ref{sec:disc} concludes with a discussion of our contributions and avenues for 
future research.  We provide reproducible code and an {\sf R} package for our 
Bayesian DGP model in a public git repository.\footnote{\url{https://github.com/stevewalsh124/dgp.hm}}
\textcolor{cyan}{Please note: this repository is currently private until we 
hear back regarding permissions for sharing the Mira-Titan data alongside our code.}

@@

\section{Mira-Titan}
\label{sec:data}

The Mira-Titan simulation suite is a dataset consisting of simulated power spectra for 
117 different cosmologies. Each of these cosmologies is specified by eight cosmological 
parameters: matter density, baryon density, amplitude of density fluctuations, 
dimensionless Hubble parameter, spectral index of scalar perturbations,
dark energy equation of state parameters, and neutrino density. For more details 
on these parameters and their effects, see \cite{dodelson2020modern, aghanim2020planck, heitmann2016mira}. 
Parameter ranges are specified in Table 2.1 of \cite{moran2023mira}.
For simplicity, we will denote these eight parameters as $\theta \in \mathbb{R}^8$. 
There is an optional ninth parameter, the redshift, which we fix at 0.
%Different simulation output exists for different values of a ninth value, 
%the redshift; in this work, we fix 
%this value at 0. 
For each cosmology, Mira-Titan contains a batch of 18 simulated spectra: an inexpensive 
spectrum estimated from linear perturbation theory ($y_p$), 16 spectra estimated from 
lower resolutions simulations ($y_{\ell_r}, r \in \{1,\dots,16\}$), and one spectrum from 
a high resolution simulation ($y_h$).

These spectra are represented as a function of wavenumber $k$.  Following 
\cite{moran2023mira}, we model on the emulation space 
$\mathcal{P}(k)=\log_{10}\left(\frac{k^{1.5}P(k)}{2\pi^2}\right)$, where $P(k)$ represents the
original spectra. Output is available for $n=351$ values of $k$, spanning
$0.001 \leq k \leq 5$, but each data type has unique wavenumber ranges
where it is deemed approximately unbiased. For $k<0.04$, only $y_p$ provides a reliable 
estimate of the true spectra. $y_{\ell_r}$ are valid for $0.04 \leq k \leq 0.25$, and 
$y_h$ is valid for $0.04 \leq k \leq 5$.  Figure \ref{fig:plot_data} shows an example 
of the output for a particular cosmology. \textcolor{red}{(Add some narration here?  Something
that doesn't repeat the caption but complements it.)}

From the multiple low- and high-resolution spectra across all cosmologies, 
\cite{moran2023mira} obtained precision estimates of these spectra across the $n$ 
wavenumber values ($p_1,\dots,p_n$) using a log-log regression model, and a multiplier 
$c$ for the increase in precision from the low- to high- resolution output. 
\textcolor{red}{(I'm missing a bit of context here - why did we need the precision
estimates in the first place?  Can we show a graph of $p_1,\dots, p_n$ to illustrate
how these change over $k$ values?)}
Given that the perturbation theory spectra is non-stochastic, we treat this as having a 
precision of $10^5$. We leverage these precision estimates within our model at each 
data type's unbiased wavenumber values to define 
$\Lambda_p = \text{diag}\left((10^5)_{k < 0.04}\right)$, 
$\Lambda_{\ell} = \text{diag}\left(16(p_1,\dots,p_n)_{0.04 \leq k < 0.25}\right)$, 
$\Lambda_h = \text{diag}\left(c(p_1,\dots,p_n)_{0.04 \leq k \leq 5}\right)$. 
\textcolor{red}{(Unless we are really short on space, could we show these in
actual "bmatrix" form?  Or visually somehow?)}
Here, each precision matrix is diagonal and the diagonal elements consist of the 
precision values or zero, depending on whether the $k$ value is approximately unbiased 
for that data type. With this, we calculate a weighted average for each cosmology: 
$\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_h y_h + \Lambda_{\ell} \bar{y}_\ell)$, 
where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$ is the precision matrix for 
the weighted average for a particular cosmology. \textcolor{red}{(Revisit the figure and
comment on the weighted average.  Explain why the weighted
average on its own is not an effective estimate.  Motivate why we need a more complex model.)}

We consider a training-testing split with 111 cosmologies used for estimation and 6 
cosmologies held out for testing.  Our ultimate objective is to effectively predict
the underlying power matter spectra for the held-out cosmologies based on their 
8-dimensional input configuration (Section \ref{sec:pred}).  But first, we must 
obtain effective estimates of the underlying spectrum for each training 
cosmology (Section \ref{sec:hm_fit}).
